---
title: "Introduction to swapper"
author: 
  - name: Milan Malfait
    affiliation:
    - Ghent University
    email: milan.malfait94@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('swapper')`"
vignette: >
  %\VignetteIndexEntry{Introduction to swapper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

The __*swapper*__ package implements a simple DE simulator based on feature
swapping.The essence of the method involves randomly scrambling a subset of
features in one subgroup of the data. This effectively induces a DE signal for
those scrambled genes, while retaining the characteristics of the original data
set without having to rely on any modeling assumptions.


# Quick start

```{r "start", message=FALSE}
library(swapper)
```


## Mock up a data set

First we mock up a dataset.

```{r sce}
sce <- scuttle::mockSCE(ncells = 100, ngenes = 50)
sce
```

This mock data set contains a `"Treatment"` variable, which is just a random partitioning of the cells in 2 groups.

```{r}
table(sce$Treatment)
```

## Simulate DE by swapping features

Next we use __*swapper*__ to simulate DE genes, using the `"Treatment"` variable
as the grouping factor. We induce DE in 10% of the original genes for cells
belonging to the `"treat1"` group.

```{r simulateDE}
set.seed(123) # for reproducibility

cells_to_swap <- sce$Treatment == "treat1"
sim <- simulateDE(sce, which_cols = cells_to_swap, prop_DE = 0.1)
rowData(sim)
table(rowData(sim)$is_DE)
```

## Explore the simulated data

From the `rowData` we can retrieve the true DE genes. The `rowData` also
contains information regarding which gene was swapped with which.

```{r trueDE_genes}
(trueDE_genes <- rownames(sim)[rowData(sim)$is_DE])
rowData(sim)[trueDE_genes, ]
```

The `"sim_group"` column in the `colData` indicates for which cells the swapping occurred. In this example, this should be equivalent to the `"Treatment"` grouping.

```{r}
colData(sim)
table(sim$Treatment, sim$sim_group)
```

We can visualize these and compare them to their original counts to see how the
swapping works.

```{r visualize, warning=FALSE, fig.width=9, fig.cap="Comparing expression values for the simulated DE genes between the original data (left) and simulated data (right)."}
if (requireNamespace("scater", quietly = TRUE)) {
    library(scater)
    
    ## Log-normalize counts for visualization
    sce <- logNormCounts(sce)
    sim <- logNormCounts(sim)
    
    p_orig <- plotExpression(
        sce, features = trueDE_genes, x = "Treatment", colour_by = "Treatment"
    ) + ggtitle("Original counts")
    
    p_sim <- scater::plotExpression(
        sim, features = trueDE_genes, x = "Treatment", colour_by = "Treatment"
    ) + ggtitle("Simulated counts")
    
    gridExtra::grid.arrange(p_orig, p_sim, ncol = 2)
}
```

From this plot and the `rowData` shown above, we can see that the counts for
e.g. `Gene_0003` in the `treat1` group of the simulated data originated from the
`treat1` counts for `Gene_0014` in the original data. Likewise, the simulated
`treat1` counts for `Gene_0014` are the original `treat1` counts for
`Gene_0042`. So we effectively induced DE by selectively swapping out counts in
one, but not the other, treatment group.


# Session Info {-}

<details><summary>Session info</summary>

```{r session_info, echo=FALSE}
## session info
options(width = 120)
sessioninfo::session_info()
```

</details>
